---
title: "hexsheff"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse); library(sf); library(geojsonio)
library(broom); library(rgeos) #TODO: rgeos being retired
library(viridis); library(scales); library(ggrepel)

# Disable use of scientific notation
options(scipen=999)
```

Resources to help create hexagonal cartograms for Sheffield.

## Hex maps vs choropleths

Choropleth maps are a great way to illustrate differences in a city. However, some of Sheffield is a part of a national park and the city's population density varies more than most cities. Typically, city sub-areas are of approximately equal population sizes, which means Sheffield sub-areas are of significantly different geographical sizes. For Sheffield choropleths, the differences in geographical sizes can detract from the main statistic of interest.

```{r choropleth, echo=FALSE}

# READ ------------------------------------

# Read the data we want to plot by hexagon
df_imd_score_by_lac <- read_rds(file = "data/df_imd_score_by_lac.rds")

# Get the LAC boundaries
sf_lac <- read_rds("data/sf_lac.rds") %>% 
  select(-objectid, -ca_name, -asc_name)

# TRANSFORM ------------------------------

# Join the SLI data to the ASC locality boundaries
sf_imd_score_by_lac <- left_join(sf_lac, df_imd_score_by_lac, 
                        by = "ca_number") %>% 
  relocate(c(X, Y), .before = geom)

# PLOT -------------------------------------

# Choropleth
ggplot(sf_imd_score_by_lac, 
       aes(fill = get('Index of Multiple Deprivation 2019 (IMD) Score'))) +
  geom_sf(alpha = 0.8, colour = 'white', size = 0.3) +
  theme_void() +
  theme(legend.position = "top",
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 8)) +
  geom_text_repel(
    size = 2.5, force = 0, bg.color = "white", bg.r = 0.15,
    aes(X, Y,
        label = 
          str_c(asc_name, "\n",
          round(get('Index of Multiple Deprivation 2019 (IMD) Score'))))) +
  scale_fill_viridis(discrete = FALSE,
                     name = "LAC IMD score (2019)",
                     direction = -1,
                     labels = function(x) round(x, digits = 0),
                     guide = guide_colourbar(
                       direction = "horizontal",
                       barheight = unit(2, units = "mm"),
                       barwidth = unit(50, units = "mm"),
                       draw.ulim = F,
                       title.position = 'top',
                       title.hjust = 0.5,
                       label.hjust = 0.5))
```

Hexagonal cartograms, or "hex maps", are one alternative to choropleths. They can still provide an indication of location, but mute differences in geographical size.

```{r hexmap, echo=FALSE}

# READ ------------------------------------

# Read the hexagon descriptions
hex <- geojson_read("data/lac-asc-hex.geojson", what="sp")

# TRANSFORM ------------------------------

# Fortify the hexagons for ggplot
hex_fortified <- tidy(hex, region = "ca_number") %>% 
  mutate(id = as.integer(id))

# Join the hexagons to the data we want to plot 
hex_fortified <- hex_fortified %>% 
  left_join(df_imd_score_by_lac, by = c("id" = "ca_number"), all.y=TRUE)

# Get hex centres and join to data we're plotting
hex_centres <- cbind.data.frame(data.frame(gCentroid(hex, byid=TRUE), 
                                       id=hex@data$ca_number)) %>%
  left_join(df_imd_score_by_lac, by=c("id"="ca_number"))

# PLOT -------------------------------------

# Hex map
ggplot() +
  geom_polygon(data = hex_fortified, aes(x = long, y = lat, group = group, 
               fill = get('Index of Multiple Deprivation 2019 (IMD) Score')),
               colour = "white") +
  coord_fixed() + 
  theme_void() +
  theme(legend.position = c(0.05, 0.7),
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 8)) +
  geom_text_repel(
    data = hex_centres, size = 3.5, force = 0, bg.color = "white", bg.r = 0.15,
    aes(x, y, 
        label = str_c(ca_name, "\n",
                      round(get('Index of Multiple Deprivation 2019 (IMD) Score'))))) +
  scale_fill_viridis(discrete = FALSE,
                     name = "LAC IMD score (2019)",
                     direction = -1,
                     labels = function(x) round(x, digits = 0),
                     guide = guide_colourbar(
                       direction = "horizontal",
                       barheight = unit(2, units = "mm"),
                       barwidth = unit(50, units = "mm"),
                       draw.ulim = F,
                       title.position = 'top',
                       title.hjust = 0.5,
                       label.hjust = 0.5))
```

ODI Leeds have done some interesting work on [hex maps](https://open-innovations.org/blog/2017-05-08-mapping-election-with-hexes).

## Building a hex map template

The process for creating the Local Area Committee hexes (which have the same boundaries as the Adult Social Care localities):

1.  Drop the `data/lac_asc_xref.csv` file into the ODI Leeds [hex map builder](https://open-innovations.org/projects/hexmaps/builder.html), hexify, and save hexes as HexJson.
2.  Drop the `data/lac_asc.hexjson` file into Oli Hawkins [HexJSON Editor](https://olihawkins.com/project/hexjson-editor/), edit, and download as GeoJson to file `data/lac-asc-hex.geojson`.

The [HexJSON format](https://open-innovations.org/projects/hexmaps/hexjson.html) defined by ODI Leeds is a great idea. However, use of the format hasn't been huge, so we've stuck with GeoJson, which plays well with the ggplot2 R package.

[geojson.io](https://geojson.io/) is useful for viewing and editing GeoJson files.
