---
title: "hexsheff"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse); library(sf); library(geojsonio)
library(broom); library(rgeos) #TODO: rgeos being retired
library(viridis); library(scales); library(ggrepel)

# Disable use of scientific notation
options(scipen=999)
```

Resources to help create hexagonal cartograms for Sheffield.

## Hex maps vs choropleths

Choropleth maps are a great way to illustrate differences in a city. However, Sheffield is part national park and population density varies more than most cities. Often, city sub-areas are of approximately equal population sizes, which means Sheffield sub-areas are of significantly different geographical sizes. For Sheffield choropleths, the differences in geographical sizes can detract attention from the main statistic of interest.

```{r choropleth, echo=FALSE}

# READ ------------------------------------

# Read the data we want to plot by hexagon
df_imd_score_by_lac <- read_rds(file = "data/df_imd_score_by_lac.rds")

# Get the LAC boundaries
sf_lac <- read_rds("data/sf_lac.rds") %>% 
  select(-objectid, -ca_name, -asc_name)

# TRANSFORM ------------------------------

# Join the SLI data to the ASC locality boundaries
sf_imd_score_by_lac <- left_join(sf_lac, df_imd_score_by_lac, 
                        by = "ca_number") %>% 
  relocate(c(X, Y), .before = geom)

# PLOT -------------------------------------

# Choropleth
ggplot(sf_imd_score_by_lac, 
       aes(fill = get('Index of Multiple Deprivation 2019 (IMD) Score'))) +
  geom_sf(alpha = 0.8, colour = 'white', size = 0.3) +
  theme_void() +
  theme(legend.position = "top",
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 8)) +
  geom_text_repel(
    size = 3.5, force = 0, bg.color = "white", bg.r = 0.15,
    aes(X, Y,
        label = 
          str_c(ca_name, "\n",
          round(get('Index of Multiple Deprivation 2019 (IMD) Score'))))) +
  scale_fill_viridis(discrete = FALSE,
                     name = "LAC IMD score (2019)",
                     direction = -1,
                     labels = function(x) round(x, digits = 0),
                     guide = guide_colourbar(
                       direction = "horizontal",
                       barheight = unit(2, units = "mm"),
                       barwidth = unit(50, units = "mm"),
                       draw.ulim = F,
                       title.position = 'top',
                       title.hjust = 0.5,
                       label.hjust = 0.5))
```

Hexagonal cartograms, or "hex maps", are one alternative to choropleths. They still provide an indication of location, but mute differences in geographical size.

```{r hexmap, echo=FALSE}

# READ ------------------------------------

# Read the hexagon descriptions
hex <- geojson_read("data/lac-asc-hex.geojson", what="sp")

# TRANSFORM ------------------------------

# Fortify the hexagons for ggplot
hex_fortified <- tidy(hex, region = "ca_number") %>% 
  mutate(id = as.integer(id))

# Join the hexagons to the data we want to plot 
hex_fortified <- hex_fortified %>% 
  left_join(df_imd_score_by_lac, by = c("id" = "ca_number"), all.y=TRUE)

# Get hex centres and join to data we're plotting
hex_centres <- cbind.data.frame(data.frame(gCentroid(hex, byid=TRUE), 
                                       id=hex@data$ca_number)) %>%
  left_join(df_imd_score_by_lac, by=c("id"="ca_number"))

# PLOT -------------------------------------

# Hex map
ggplot() +
  geom_polygon(data = hex_fortified, aes(x = long, y = lat, group = group, 
               fill = get('Index of Multiple Deprivation 2019 (IMD) Score')),
               alpha = 0.8, colour = "white", size = 0.3) +
  coord_fixed() + 
  theme_void() +
  theme(legend.position = c(0.05, 0.7),
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 8)) +
  geom_text_repel(
    data = hex_centres, size = 3.5, force = 0, bg.color = "white", bg.r = 0.15,
    aes(x, y, 
        label = str_c(ca_name, "\n",
                      round(get('Index of Multiple Deprivation 2019 (IMD) Score'))))) +
  scale_fill_viridis(discrete = FALSE,
                     name = "LAC IMD score (2019)",
                     direction = -1,
                     labels = function(x) round(x, digits = 0),
                     guide = guide_colourbar(
                       direction = "horizontal",
                       barheight = unit(2, units = "mm"),
                       barwidth = unit(50, units = "mm"),
                       draw.ulim = F,
                       title.position = 'top',
                       title.hjust = 0.5,
                       label.hjust = 0.5))
```

ODI Leeds have done some interesting work on [hex maps](https://open-innovations.org/blog/2017-05-08-mapping-election-with-hexes).

## Building a hex map template

Once a template for Sheffield sub-area hexagons is defined, e.g. by a GeoJson file, it can be reused to illustrate different statistics. The process for creating the LAC (Local Area Committee) hexagons was:

1.  Drop the `data/lac_asc_xref.csv` file into the ODI Leeds [hex map builder](https://open-innovations.org/projects/hexmaps/builder.html), hexify, and save hexes as HexJson.
2.  Drop the `data/lac_asc.hexjson` file into Oli Hawkins [HexJSON Editor](https://olihawkins.com/project/hexjson-editor/), edit, and download as GeoJson to file `data/lac-asc-hex.geojson`.

The [HexJSON format](https://open-innovations.org/projects/hexmaps/hexjson.html) defined by ODI Leeds is a great idea. However, use of the format hasn't been huge, so we've stuck with GeoJson, which plays well with the ggplot2 R package.

[geojson.io](https://geojson.io/) is useful for viewing and editing GeoJson files.

## Different Sheffield hex maps  

So far, this repository includes a hex map for LACs: `data/lac-asc-hex.geojson`

ASC (Adult Social Care) localities have the same boundaries as LACs. The LAC hex map includes ASC locality cross-references, so it can be used for that purpose too.

If hex maps prove useful, other Sheffield hex maps that could be created include:  

1.  Wards (28)  
2.  Neighbourhoods (100)  

The more the audience of Sheffield hex maps use them, the easier they will be to read. Ideally, data analysts producing the hex maps will agree on a single hex map for each type of sub-area. New or improved Sheffield hex maps, or suggestions, are encouraged via pull requests or issues.
